---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
pp_stan <- identical(Sys.getenv("PP_README_STAN"), "1") &&
  requireNamespace("cmdstanr", quietly = TRUE)
```

# pseudopost

**Post-hoc pseudo-posterior uncertainty adjustment for composite and
pseudo-likelihood Stan models.**

Composite or pairwise likelihoods can count dependent terms from the same
respondent as though they were independent. The resulting pseudo-posterior
spread may misrepresent the repeated-sampling uncertainty of estimates:
intervals can be too narrow or too wide. `pseudopost` adjusts draws from
[cmdstanr](https://mc-stan.org/cmdstanr/) fits toward a robust covariance;
finite-sample coverage is not guaranteed.

## What it does

`adjust_pseudo_posterior()` offers:

- **IJ** (`estimator = "ij"`): a draw-based infinitesimal jackknife covariance
  target using per-person log-likelihood contributions (Giordano & Broderick,
  2023).
- **Closed sandwich** (`"sandwich"`): calibrates unconstrained draws to the
  Godambe covariance $V = Q^{-1} P Q^{-1}$, where $Q$ is curvature and $P$
  is between-person score variability.
- **Open-faced sandwich** (`"ofs"`): applies a square-root transform directly
  to draws using the same $Q$ and $P$ (Shaby, 2014).

For sandwich and OFS, `bread` selects `"mcmc"`, `"hessian_mle"`, or
`"hessian_map"`; `meat` selects `"score"` or `"score_avg"`. Any estimator
can use `ci = "quantile"` or `"wald"`. These arguments accept vectors;
`bread` and `meat` are ignored for IJ. The unadjusted baseline is
`methods = "noadj"`.

`magnitude_adjust()` and `refit_magnitude_adjust()` instead apply a single
scalar to the composite likelihood; a refit is needed, and the scalar cannot
generally reproduce direction-dependent sandwich uncertainty.

See the [model requirements vignette](vignettes/pseudo-likelihood-requirements.Rmd)
for the Stan contract, distinctions between methods, and a worked example.

## Installation

```{r eval = FALSE}
# From GitHub
remotes::install_github("marklhc/pseudopost")
```

Sampling and Hessian-based adjustments require `cmdstanr` and a working
[CmdStan](https://mc-stan.org/cmdstan/) toolchain.

## Quick start

The shipped normal model is a simple check: under correct specification,
adjustments should be close to the unadjusted result in large samples.

```{r, eval = pp_stan}
library(cmdstanr)
library(pseudopost)

model <- cmdstan_model(
  system.file("test-models", "truelik.stan", package = "pseudopost"),
  force_recompile = TRUE, quiet = TRUE)
set.seed(42)
dat <- list(N = 100L, y = rnorm(100, 1, 1),
            magnitude_adj = 1.0, use_priors = 1L)
fit <- model$sample(data = dat, chains = 2L, iter_warmup = 200L,
                    iter_sampling = 200L, seed = 42L, refresh = 0,
                    show_messages = FALSE)

check_pl_requirements(fit)
pp <- adjust_pseudo_posterior(
  fit, params = "theta",
  estimator = c("ij", "sandwich", "ofs"),
  bread = "mcmc", meat = "score", ci = "quantile",
  verbose = FALSE)
pp$table
pp$meta$warnings
```

For your model, generate a vector `log_lik` with **one entry per independent
unit** (e.g., sum all item-pair terms for each respondent). The package checks
its shape, not independence or consistency with the fitted likelihood.
`pp$table` contains estimates, SEs, and intervals; `pp$draws` contains
adjusted draws. Check `pp$meta$warnings` for method-specific failures.
Intervals default to 90% (`ci_level = 0.90`). Hessian breads require
`model` and `data` (or optimized points); see the vignette for details.

## References

- Shaby, B. A. (2014). The Open-Faced Sandwich Adjustment for MCMC Using
  Estimating Functions. *Journal of Computational and Graphical Statistics*,
  23(3), 853-876. <doi:10.1080/10618600.2013.842174>.
- Giordano, R., & Broderick, T. (2023). The Bayesian Infinitesimal Jackknife
  for Variance. arXiv preprint arXiv:2305.06466. <doi:10.48550/arXiv.2305.06466>.

## License

GPL (>= 3)
