---
title: "Model requirements for pseudo-posterior adjustment"
package: pseudopost
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model requirements for pseudo-posterior adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Register package methods for the examples.
suppressMessages(library(pseudopost))

# Opt in to compilation and sampling with PP_EXAMPLES_STAN=1.
pp_stan <- identical(Sys.getenv("PP_EXAMPLES_STAN"), "1") &&
  requireNamespace("cmdstanr", quietly = TRUE)
```

## Why adjust a pseudo-posterior?

In pairwise likelihoods, several dependent item-pair terms come from the same
respondent. Treating their product as an ordinary likelihood can produce
pseudo-posterior intervals that are too narrow or too wide for repeated-sampling
inference.

Write the composite log-likelihood as
$\ell_c(\theta) = \sum_{r=1}^n \ell_r(\theta)$, where $r$ indexes
**independent observational units** (people in the examples below). Each
$\ell_r$ may itself sum several conditional or pairwise log-probabilities.
Let $Q$ denote the curvature (the negative Hessian of the *total*
$\ell_c$), and write $s_r = \nabla\ell_r$ for respondent $r$'s score.
The score variability is estimated by
$P = \sum_{r=1}^n (s_r - \bar s)(s_r - \bar s)^{\mathsf T}$.
The sandwich covariance is $V = Q^{-1} P Q^{-1}$ (Godambe & Heyde, 1975).
$Q$ and $P$ are the *bread* and *meat*, also denoted $H$ and $J$.
With `bread = "mcmc"`, $Q$ is estimated as the inverse covariance of the
unconstrained draws rather than a Hessian. These adjustments approximate
sampling uncertainty without refitting MCMC; they do not guarantee
finite-sample coverage.

`adjust_pseudo_posterior()` offers three estimators:

- **Infinitesimal jackknife** (`estimator = "ij"`): estimates a covariance
  target from the MCMC draws and per-unit log-likelihood contributions,
  without explicitly computing a Hessian or per-unit score derivatives
  (Giordano & Broderick, 2023). This interface uses the draw-based proxy IJ.
- **Closed Godambe sandwich** (`estimator = "sandwich"`): estimates $Q$ and $P$
  separately, then transforms the draws to have covariance $V$ on the
  unconstrained parameter scale.
- **Open-faced sandwich (OFS)** (`estimator = "ofs"`): estimates $Q$ and $P$
  separately and applies $\Omega = (Q^{-1}P)^{1/2}$ directly to centered draws
  (Shaby, 2014), without first matching their empirical covariance to $V$.

OFS gives covariance $\Omega C_{\mathrm{emp}}\Omega^{\mathsf T}$, where
$C_{\mathrm{emp}}$ is the covariance of the original unconstrained draws.
It equals $V$ if $C_{\mathrm{emp}} = Q^{-1}$. With `bread = "mcmc"`, this
holds by construction **in any dimension**: OFS and closed sandwich have the
same unconstrained covariance, though transformed draws and nonlinear-parameter
intervals can differ. With a Hessian bread, their covariances may differ.

The `estimator`, `bread`, `meat`, and `ci` arguments accept vectors of choices.
`bread` and `meat` do not apply to IJ. `methods = "noadj"` gives the unadjusted
baseline; other legacy `methods` presets remain available.

| argument    | values                                     | meaning |
|-------------|--------------------------------------------|---------|
| `estimator` | `"ij"`, `"sandwich"`, `"ofs"`              | draw-based IJ, closed sandwich, or OFS |
| `bread`     | `"mcmc"`, `"hessian_mle"`, `"hessian_map"` | inverse draw covariance or Hessian-based curvature $Q$; sandwich/OFS only |
| `meat`      | `"score"`, `"score_avg"`                   | per-unit score covariance $P$ at one point or averaged over draws; sandwich/OFS only |
| `ci`        | `"quantile"`, `"wald"`                     | adjusted-draw percentiles or mean $\pm z$ times the adjusted-draw SD |

A scalar **magnitude adjustment** (`magnitude_adjust()` and
`refit_magnitude_adjust()`) rescales the whole composite likelihood and needs
a refit. It cannot generally reproduce direction-dependent sandwich covariance.

## What a qualifying model must expose

A fit must satisfy two basic requirements for post-hoc adjustment:

1. **It is a cmdstanr fit**---a `CmdStanMCMC` (or `CmdStanMLE`) object; and
2. **Its Stan program generates per-unit log-likelihood contributions**---a
   vector named `log_lik` by default, with one entry for each independent
   person or other observational unit. For pairwise likelihoods, sum *all*
   item-pair contributions for one respondent into that respondent's entry;
   do not use one entry per item pair. The score covariance uses derivatives
   of these entries, while the IJ uses their variation across draws.
   `check_pl_requirements()` checks its vector shape, not independence or
   agreement with the fitted composite likelihood.

For a scalar magnitude adjustment, the Stan program must also declare the
`magnitude_adj` data argument and use it to multiply the entire composite
log-likelihood. `refit_magnitude_adjust()` checks that the **refit model**
declares the argument, but cannot verify that the code scales every contribution.

Two further requirements apply to the Hessian-based breads
(`"hessian_mle"` / `"hessian_map"`):

- They require `model` and `data` or a pre-optimized `opt` / `opt_map` to
  evaluate the Hessian at an MLE or MAP, not the MCMC posterior mean.
- For `"hessian_mle"` to be a true likelihood-only MLE, the model must
  disable **all** prior terms when `use_priors = 0`; the package cannot
  verify that the Stan program honors this switch. Check optimizer
  convergence and Hessian conditioning as well: an optimizer exit code of
  zero does not ensure a well-conditioned covariance estimate.

### A minimal qualifying model

The shipped `inst/test-models/truelik.stan` shows the required structure
for independent normal observations:

```stan
data {
  int<lower=1> N;
  vector[N] y;
  real<lower=0> magnitude_adj;   // needed for a magnitude refit
  int<lower=0, upper=1> use_priors;
}
parameters { real theta; }
model {
  if (use_priors == 1) theta ~ normal(0, 10);
  for (n in 1:N) target += magnitude_adj * normal_lpdf(y[n] | theta, 1);
}
generated quantities {
  vector[N] log_lik;             // one entry per independent observation
  for (n in 1:N) log_lik[n] = normal_lpdf(y[n] | theta, 1);
}
```

For an Ising pseudo-likelihood or pairwise factor model, sum each person's
conditional or pairwise terms in **one entry per person**, not per symptom
or item pair. At unit magnitude, `sum(log_lik)` should equal the composite
log-likelihood in the model block, excluding priors.

## Checking the requirements

`check_pl_requirements()` returns, invisibly, a list with `m` (the number
of entries in `log_lik`) and `ok = TRUE`. A missing `log_lik` or a
matrix-valued `log_lik` is rejected. A single entry is accepted with a
warning because a per-unit covariance cannot be meaningfully estimated
from just one unit. Setting `need_magnitude = TRUE` additionally checks
that the fit's Stan program declares `magnitude_adj`.

```{r, eval = pp_stan}
library(cmdstanr)
model <- cmdstan_model(
  system.file("test-models", "truelik.stan", package = "pseudopost"),
  force_recompile = TRUE, quiet = TRUE)
set.seed(42)
dat <- list(N = 100L, y = rnorm(100, 1, 1),
            magnitude_adj = 1.0, use_priors = 1L)
fit <- model$sample(data = dat, chains = 2L, iter_warmup = 200L,
                    iter_sampling = 200L, seed = 42L, refresh = 0,
                    show_messages = FALSE)
res <- check_pl_requirements(fit)         # returns invisibly
res$m                                     # 100
check_pl_requirements(fit, need_magnitude = TRUE)
```

## Running an adjustment

`adjust_pseudo_posterior()` returns adjusted draws and estimates, SEs, and
intervals for each requested combination and parameter. `ci = "quantile"`
uses adjusted-draw percentiles; `"wald"` uses mean $\pm z$ times SD.
Both default to 90% intervals (`ci_level = 0.90`); observed repeated-sampling
coverage may differ.

```{r, eval = pp_stan}
pp <- adjust_pseudo_posterior(
  fit, "theta",
  estimator = c("ij", "sandwich", "ofs"),
  bread = c("mcmc", "hessian_map"),
  meat = "score",
  ci = "quantile",
  model = model, data = dat, init = list(list(theta = 0.0)), verbose = FALSE)
pp$table
pp$meta$warnings
```

Under the correctly specified normal model, $P = Q$ in expectation, so
adjustments should approach the unadjusted result with enough data. For
composite likelihoods or misspecified models, adjusted SEs can be **larger
or smaller**. Assess coverage across repeated samples, not one dataset.

## The `pseudo_post` result object

The `pseudo_post` object has `table` (`method`, `param`, `est`, `se`, `lo`,
`hi`), `draws` (by method), and `meta` (settings and diagnostics); it supports
`print()`, `summary()`, and `as.data.frame()`. Check `meta$warnings`: one
method can fail while others return results.

## References

- Shaby, B. A. (2014). The Open-Faced Sandwich Adjustment for MCMC Using
  Estimating Functions. *Journal of Computational and Graphical Statistics*,
  23(3), 853-876. <doi:10.1080/10618600.2013.842174>.
- Giordano, R., & Broderick, T. (2023). The Bayesian Infinitesimal Jackknife
  for Variance. arXiv preprint arXiv:2305.06466. <doi:10.48550/arXiv.2305.06466>.
- Godambe, V. P., & Heyde, C. C. (1975). Quasi-likelihood and optimal
  estimation. *Journal of the American Statistical Association*, 70(351),
  455-467.
