---
title: "The pseudo-likelihood requirements"
package: pseudopost
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The pseudo-likelihood requirements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Load the package so its S3 methods (print/summary/as.data.frame.pseudo_post)
# are registered for every chunk below. R CMD check installs the package into
# the vignette library, so this is safe in all environments.
suppressMessages(library(pseudopost))

# The Stan-backed chunks (compile + sample) run only when opted in, so this
# vignette builds in any environment. Set PP_EXAMPLES_STAN=1 and install the
# CmdStan toolchain to run them.
pp_stan <- identical(Sys.getenv("PP_EXAMPLES_STAN"), "1") &&
  requireNamespace("cmdstanr", quietly = TRUE)
```

## Why adjust the pseudo-posterior at all?

When a Stan model is fit to a *composite*, *pairwise*, or otherwise
*pseudo-likelihood* target, the function being sampled (treated as if it were
a likelihood) is **not** a true joint log-likelihood. The posterior that MCMC
recovers is therefore the posterior of the pseudo-likelihood, and the
**naive** posterior covariance
(the inverse Hessian, or simply the sample covariance of the draws) is
generally *miscalibrated*: it is often, but not always, *liberal*
(understating the true variability, so that standard errors/uncertainty estimates are too small and
credible intervals too narrow).

`pseudopost` provides post-hoc covariance (sandwich) adjustments that target
the robust covariance $V = H^{-1} J H^{-1}$ (Shaby, 2014) to improve scale and
coverage without refitting the model---all exposed through a single entry
point, `adjust_pseudo_posterior()`. This is an approximation to the correct
uncertainty, not a guarantee of correct coverage:

- an **infinitesimal jackknife** (`estimator = "ij"`) estimator — a
  model-blind estimate of the full $V$ straight from the MCMC draws
  (Giordano & Broderick, 2023);
- a **closed** (`estimator = "sandwich"`) Godambe estimator with an explicit
  precision `bread` ($H$) and score `meat` ($J$);
- an **open-faced** (`estimator = "ofs"`) sandwich estimator (Shaby, 2014):
  the same explicit `bread` ($Q$) and `meat` ($P$), applied as the draw
  transform $\Omega = (Q^{-1} P)^{1/2}$ without calibrating to the empirical draw
  covariance;
- a `ci` type (`"quantile"` or `"wald"`).

The `"ij"` estimator is the infinitesimal jackknife (IJ) of Giordano &
Broderick (2023): it estimates the full $V$ directly from the draws without
estimating $H$ or $J$ separately. It is related to, but not the same as,
Shaby's (2014) "open-faced sandwich": that term denotes the half-sandwich
$JH^{-1}$, which Shaby's adjustment joins to the quasi-posterior's own
one-bread covariance $H^{-1}$ to complete $V$.

The open-faced (`estimator = "ofs"`) estimator is the third option and
Shaby's (2014) original adjustment. It takes the draws' own covariance
$C_{\text{emp}}$ to be the bread $Q^{-1}$ (the "one-bread" asymptotic) and applies the
draw transform $\Omega = (Q^{-1} P)^{1/2}$ directly to the centered draws,
without measuring or inverting $C_{\text{emp}}$. The closed sandwich instead estimates
the full $V = H^{-1} J H^{-1}$ and calibrates the adjusted draws to that target,
so their (unconstrained) covariance equals $V$ by construction; the
infinitesimal jackknife is model-blind, estimating $V$ from the draws alone.
OFS and the closed sandwich agree when $C_{\text{emp}} = Q^{-1}$ (e.g. with
`bread = "mcmc"` in one dimension, where $Q^{-1}$ is by construction the
empirical draw covariance).

The four axes (`estimator`, `bread`, `meat`, `ci`) are orthogonal and each
may be a vector, so a call runs a sub-grid with one row per parameter and
valid combination. The `"ij"` estimator does not use `bread`/`meat`; the
baseline (no robust covariance) is the legacy `methods = "noadj"` preset, and
the original named presets remain as silent aliases for the axes.

| axis        | values                                        | role |
|-------------|-----------------------------------------------|------|
| `estimator` | `"ij"`, `"sandwich"`, `"ofs"`                 | infinitesimal jackknife (model-blind $V$ from the draws), closed Godambe $H^{-1} J H^{-1}$, or open-faced draw transform $(Q^{-1} P)^{1/2}$ |
| `bread`     | `"mcmc"`, `"hessian_mle"`, `"hessian_map"`    | the precision $H$; `"sandwich"`/`"ofs"` only |
| `meat`      | `"score"`, `"score_avg"`                      | the score moment $J$; `"sandwich"`/`"ofs"` only |
| `ci`        | `"quantile"`, `"wald"`                        | percentile of the adjusted draws vs estimate $\pm z \cdot \text{se}$ |

A scalar **magnitude** (scale) adjustment is available via
`magnitude_adjust()` / `refit_magnitude_adjust()`. It is a single *global*
scalar that rescales the whole pseudo-likelihood---a coarse correction of the
overall posterior scale that, in general, does not reproduce a
direction-dependent multivariate sandwich covariance.

The rest of this vignette explains the **requirements** a model must satisfy
for any of these adjustments to be available.

## What a qualifying model must expose

A model is eligible for post-hoc adjustment if and only if:

1. **It is a cmdstanr fit**---a `CmdStanMCMC` (or `CmdStanMLE`) object; and
2. **It exposes a per-observation log-likelihood generated quantity**---a
   vector named (by default) `log_lik`, with one entry per pseudo-observation.
   This vector is what the adjustments are built from: the sandwich `meat` is
   its per-observation Jacobian, and the IJ uses it directly.
   `check_pl_requirements()` validates the vector **shape** only (that
   `log_lik` exists and is a vector of length `m`); it does **not** verify
   that the entries are contributions to independent observational units or
   that they sum to the fitted pseudo-likelihood---that is the model author's
   responsibility.

If you also want the scalar *magnitude* (scale) adjustment, the model must
**additionally accept a `magnitude_adj` data argument**, which rescales the
whole pseudo-likelihood (`target += magnitude_adj * <pseudo-likelihood>`).

Two further requirements apply to the Hessian-based breads
(`"hessian_mle"` / `"hessian_map"`):

- They evaluate the Hessian at an MLE/MAP, so they require a `model`/`data`
  pair (or a pre-optimized `opt`/`opt_map`); currently there is **no** option for using MCMC-posterior-mean (the posterior mean is not an MLE or MAP and
  would mislabel the point).
- For `"hessian_mle"` to be a true likelihood-only MLE, the model must
  disable **all** priors when `use_priors = 0`---the package cannot verify
  that the Stan program honours that switch.

### A minimal qualifying model

The package ships a tiny one-parameter model, `truelik.stan`, in
`inst/test-models/`. It is a *true* per-observation likelihood, which is
useful as a sanity check: every correction must be close to the identity function.
Its essential structure is:

```stan
data {
  int<lower=1> N;
  vector[N] y;
  real<lower=0> magnitude_adj;   // <- required for the magnitude adjustment
  int<lower=0, upper=1> use_priors;
}
parameters { real theta; }
model {
  if (use_priors == 1) theta ~ normal(0, 10);
  for (n in 1:N) target += magnitude_adj * normal_lpdf(y[n] | theta, 1);
}
generated quantities {
  vector[N] log_lik;             // <- required: one entry per pseudo-observation
  for (n in 1:N) log_lik[n] = normal_lpdf(y[n] | theta, 1);
}
```

A genuine *pseudo*-likelihood model has the same shape, except the per-
observation terms in the `model` block are pairwise / composite contributions
rather than a joint likelihood. The requirements are unchanged: expose
`log_lik` (one entry per pseudo-observation) and, for the magnitude
adjustment, accept `magnitude_adj`.

## Checking the requirements

`check_pl_requirements()` validates the requirements and returns, invisibly,
the number of pseudo-observations `m` (the length of `log_lik`). If
`log_lik` is missing
it stops with a clear error naming the generated quantity it could not find;
a **matrix-valued** `log_lik` (e.g. stored with 2-D brackets such as
`log_lik[i, j]`) is also rejected, since the requirements call for a
**vector** with one entry per pseudo-observation; with `need_magnitude = TRUE`
it additionally requires the `magnitude_adj` data argument.

```{r, eval = pp_stan}
library(cmdstanr)
model <- cmdstan_model(
  system.file("test-models", "truelik.stan", package = "pseudopost"),
  force_recompile = TRUE, quiet = TRUE)
set.seed(42)
dat <- list(N = 100L, y = rnorm(100, 1, 1),
            magnitude_adj = 1.0, use_priors = 1L)
fit <- model$sample(data = dat, chains = 2L, iter_warmup = 200L,
                    iter_sampling = 200L, seed = 42L, refresh = 0,
                    show_messages = FALSE)
res <- check_pl_requirements(fit)         # returns invisibly
res$m                                     # 100
check_pl_requirements(fit, need_magnitude = TRUE)
```

## Running an adjustment

`adjust_pseudo_posterior()` runs one or more adjustments across the
`estimator` / `bread` / `meat` / `ci` axes on a fit that meets the
requirements and
returns a `pseudo_post` object containing a long-form table
(`method, param, est, se, lo, hi`), the adjusted draws per method, and run
metadata. Each axis may be a vector, so the axes run a sub-grid (one row per
valid combination and parameter).

```{r, eval = pp_stan}
# New interface: the systematic estimator / bread / meat / ci grid.
pp <- adjust_pseudo_posterior(
  fit, "theta",
  estimator = "sandwich",
  bread = c("mcmc", "hessian_map"),
  meat = "score",
  ci = "quantile",
  model = model, data = dat, init = list(list(theta = 0.0)), verbose = FALSE)
print(pp)
```

The original named presets remain as silent aliases for these axes (`methods =
"noadj"` is the unadjusted baseline):

```{r, eval = pp_stan}
pp_legacy <- adjust_pseudo_posterior(
  fit, "theta",
  methods = c("noadj", "ijse", "ijse2", "curvadj", "curvadj2", "curvadj3"),
  model = model, data = dat, init = list(list(theta = 0.0)), verbose = FALSE)
print(pp_legacy)
```

Because `truelik` is a *true* likelihood, the information equality makes the
meat equal the bread, so every correction is ~ the identity: the adjusted
standard errors (e.g. the `"ij"` and `"sandwich"` rows) land within a few
percent of the unadjusted `"noadj"` one. For a genuine pseudo-likelihood the
adjusted standard errors are typically *larger* than the naive ones —
precisely the anti-conservatism being corrected.

## The `pseudo_post` result object

The result is a plain list of class `pseudo_post` with three elements:
`table`, `draws`, and `meta`. It supports `print()`, `summary()`, and
`as.data.frame()`:

```{r}
set.seed(1)
pp <- structure(
  list(table = data.frame(method = c("noadj", "ij.quant"),
                          param = "theta",
                          est = c(1.03, 1.05), se = c(0.09, 0.14),
                          lo = c(0.86, 0.77), hi = c(1.20, 1.33),
                          stringsAsFactors = FALSE),
       draws = list(noadj = matrix(rnorm(20), 10, 2),
                    ij.quant = matrix(rnorm(20), 10, 2)),
       meta = list(fit = NULL, methods = c("noadj", "ij.quant"), ci_level = 0.90,
                    ij_method = "proxy", warnings = character(0),
                   notes = character(0))),
  class = "pseudo_post")
print(pp)
as.data.frame(pp)
```

## References

- Shaby, B. A. (2014). The Open-Faced Sandwich Adjustment for MCMC Using
  Estimating Functions. *Journal of Computational and Graphical Statistics*,
  23(3), 853-876. <doi:10.1080/10618600.2013.842174>.
- Giordano, R., & Broderick, T. (2023). The Bayesian Infinitesimal Jackknife
  for Variance. arXiv preprint arXiv:2305.06466. <doi:10.48550/arXiv.2305.06466>.
- Ji, F., Lee, J., & Rabe-Hesketh, S. (2025). Valid standard errors for
  Bayesian quantile regression with clustered and independent data.
  *Journal of Educational and Behavioral Statistics*, 51(5), 1055-1084.
  <doi:10.3102/10769986251379738>.
- Godambe, V. P., & Heyde, C. C. (1975). Quasi-likelihood and optimal
  estimation. *Journal of the American Statistical Association*, 70(351),
  455-467.
